The Heavy Photon Search Test Detector

### 2013-11-12 hps-testrun-nim.tex
John's changes to mostly detector overview and intro.

%% main text
\section{Introduction}
\label{introduction}
One of the more striking anomalies in recent astrophysical observations is an excess of high energy electrons 
and positrons consistent with a large, unknown source of pair 
production~\cite{Adriani:2008zr,FermiLAT:2011ab}. 
The existence of a massive, hidden sector vector boson; a ``dark photon;" that mixes kinetically with 
the Standard Model photon and mediates annihilation or decay of dark matter particles is a theoretically 
attractive explanation of this cosmic-ray excess~\cite{Holdom:1985ag}\cite{ArkaniHamed:2008qn}.
\begin{center}
{\small
\begin{figure*}[ht]
    \includegraphics[width=0.8\textwidth]{figures/hps_testrun_rendering}
\caption{Rendering of the HPS Test apparatus installed on the beam line.}
\label{fig:testrundetector}
\end{figure*}
}
\end{center}

Heavy photons, aka known as "hidden sector" or "dark" photons, are particles with mass 10-1000 MeV which 
couple
weakly to the SM photon, so can be radiated by electrons, and can decay into e+e- pairs, albeit at rates far
below those of QED trident processes. They have been suggested by numerous Beyond Standard Model 
theories and
as an explanation of the current muonic g-2 anomaly, and may couple directly
to hidden sector particles that may constitute all or some of the dark matter. Current phenomenology highlights 
the 
20-1000 MeV mass range, and suggests that heavy photons couple to electric charge with strength 
$\epsilon$e, where 
$\epsilon$ is in the range of $10^{-3} -10^{-5}$. This range of parameters makes heavy photon searches viable 
at electron-positron colliders and in moderate energy fixed target electroproduction, and results from both
approaches have already been reported. Fixed target experiments offer the highest sensitivity 
for heavy photons with mass $<1000$~MeV~\cite{Bjorken:2009mm}, but require large data sets and good 
mass resolution to identify a small mass peak above the copious QED background. At small couplings, heavy 
photons become long-lived, so detection of a secondary decay vertex can reject the prompt QED background 
and boost experimental sensitivity.
The HPS experiment uses both invariant mass and vertex signatures by placing a compact silicon tracking and 
vertexing detector ($\approx 1$~m long) immediately downstream ($\approx 10$~cm) of a thin target 
($\approx0.25~X_0$ tungsten) and inside of a dipole magnet ($\approx 1$~T)~\cite{HPS_proposal_2010}. 
The experiment runs at high rates, exploiting the 100\% duty cycle of the 
CEBAF accelerator to accumulate the need statistics. The trigger is provided by a fast electromagnetic 
calorimeter just downstream of the magnet.

The HPS Test run, a simplified version of the full HPS experiment, was proposed and approved at JLab 
as the first stage of the HPS experiment.  Its purposes included demonstrating that the apparatus and data 
acquisition systems are technically feasible and
that the trigger rates and occupancies encountered in electron-beam running are as simulated. Additionally,
given dedicated running time with electron beams, the HPS Test Run could provide new sensitivity to heavy 
photons. The HPS Test apparatus was installed on April 19, 2012, and 
ran parasitically with the HDice experiment, using its photon beam, until May 18. The JLab run schedule 
precluded any dedicated electron beam running, but the HPS Test Run was allowed a short and valuable
dedicated run with the photon beam.

This paper reviews the HPS Test Run apparatus. It documents the successful performance of the trigger, data 
acquisition, silicon tracker, and ECal and
shows that the performance assumed in calculating the physics reach of the experiment is realistic.
Of particular importance, data from the dedicated photon beam running has been used to compare
the measured trigger rates with those expected in simulation. The Test Run trigger rate is almost entirely due to
photons which have converted to \ee{} pairs in a thin target upstream of HPS and it is sensitive to the particulars 
of their multiple Coulomb scattering in the conversion target.  Since, during electron beam running, the tails of 
the multiply Coulomb scattered primary beam are the dominant source of occupancy in the tracker and the 
trigger rate in the Ecal, they must be understood quantitatively. The trigger rates in the Test Run are also 
sensitive to the tails of the multiple Coulomb scattering distribution. Good agreement between the measured 
and simulated rates confirms our understanding of multiple Coulomb scattering and 
the reliability of the background simulation used to benchmark the physics reach of the HPS experiment.

The very small signal rate compared to the large background and the need to reconstruct the final states of 
the electron and positron in a dense radiation environment millimeters from a high-current beam 
place stringent requirements on the detector, which should have,
\begin{itemize}
\item large and uniform acceptance in the forward region to catch boosted decay products,
\item a flexible, redundant and efficient trigger selecting electron and positron pairs at rates up to 50kHz,
\item excellent track reconstruction efficiency for electrons and positrons,
\item good angular and momentum resolution to reconstruct invariant mass precisely,
\item excellent vertex resolution to search for displaced heavy photon decays,
\item high rate electronics with event timing resolution down to 2~ns,
\item data acquisition system capable of handling data rates of 100~MB/s to permanent storage,
\item detector components that can survive and efficiently operate in a high radiation environment with 
local doses exceeding $100$~Mrad
\end{itemize}

\section{Detector Overview}
\label{detector}
The HPS experiment is proposed to run in Hall B using electron beams of Continuos Electron Beam 
Accelerator Facility (CEBAF) at JLab. CEBAF can simultaneously deliver  electron 
beams to multiple experimental hall at 499 MHz frequency with beam currents of up to 150 $\mu$A (Hall-B 
is limited to $<1 ~\mu$A).  Currently CEBAF facilities are undergoing energy an upgrade that is expected to be 
completed in end of 2014. The upgraded CEBAF machine will double its maximum energy from $6$~GeV 
during the HPS test run to $12$~GeV.

The HPS Test apparatus was designed as a 
first step of a multi-stage approach 
to the HPS experiment. The primary goal was to demonstrate that the technology and design approach 
chosen are sound for running in such a challenging environment. Nevertheless, a secondary but 
important aspect was that, after proving the technical concepts,  the apparatus would have the 
capability to discover \Aprime{} events. The implication was that the design of the HPS Test apparatus 
would be subject to almost the same constraints and requirements as the full experiment.
The kinematics of \Aprime{} production typically results in final state particles within a few degrees of the 
incoming beam, especially at low $m_{\Aprime{}}$.  
%Step
This implies that the apparatus for an \Aprime{} search in electron interactions with fix targets must operate in 
close proximity to an intense electron beam that passes through a high-Z target. 
Hence it should cope with high backgrounds and tolerate the high radiation environment. 
To meet this challenge, HPS will deploy high rate and radiation hard detector composed of 
a silicon vertex tracker with 40~MHz readout and a PbWO$_4$ crystal calorimeter with 250~MHz 
readout, capable of triggering multi-prong events at top to 50~kHz rate.

%\subsection{HPS Test setup}

The HPS Test apparatus, a simplified version of that 
planned for the full HPS experiment, demonstrates the feasibility of the detector
technologies proposed for the silicon tracker, electromagnetic calorimeter, and data acquisition system. 
A rendering of the HPS Test setup is shown in Fig.~\ref{fig:testrundetector}. The setup is based on the Hall-B 
pair spectrometer (PS) magnet, 18D36 dipole. The silicon vertex tracker (SVT) was installed inside the vacuum 
chamber of PS and the electromagnetic calorimeter (ECal) was mounted behind the vacuum flange. 
A new vacuum box 
attached to the upstream end of the pair spectrometer vacuum chamber provided necessary services to 
the SVT (cooling, power and signal readout). 
Most technologies used in the HPS Test apparatus as well as the magnet and the vacuum chamber will be 
used in the HPS detector with some 
improvements to each of the systems; both from lessons learned during running of the HPS Test but 
also from planned upgrades for longer and more effective running.

%\begin{multicols}{2}
\begin{center}
\begin{table*}[t]
{\small
\caption{Overview of the coverage, segmentation and performance of the HPS Test detector. }
\begin{tabular}{lccccccc}
\hline 
System & Coverage & \# channels & ADC & Time resolution & \# layers & Segmentation & Performance \\
 & (mrad) &  & (bit) & (ns) &  &  &  \\
\hline
SVT & $15<\theta_{y} < XXXX$ & 12800 & 14 & $\approx 2$~ns & 5  & $\approx 120~\mu$m $r-\phi$ & $\sigma_{d0,y}  \approx 100~\mu$m \\
& &  &  &  & (stereo layers) & $\approx 6~\mu$m $z$ & $\sigma_{d0,x} \approx 300~\mu$m \\
& &  &  &  &  &  & $\sigma_{d0,z}\approx 1$~mm \\
\hline
ECal & $15<\theta_{y} < XXXX$ & 442 & 12 & 4~ns & 1 & $1.3\times1.3$~cm$^2$  & $\sigma(E)/E \approx 4.5\%$ \\ 
 &  &  &  &  &  & $1.6\times1.6$~cm$^2$  &  \\ 
\hline
\end{tabular}
\label{tab:detector-overview}
}
\end{table*}
\end{center}
%\end{multicols}

An overview of the coverage, segmentation and performance of the HPS Test detector is given in 
Tab.~\ref{tab:detector-overview}.  In order to accommodate the dead zone, the SVT and ECal are built in two 
halves that are mirror reflections of one 
another about the plane of the nominal electron beam.  Each half of the SVT consists of five double-sided 
modules 
mounted on a support plate that provides services to the modules and allows them to be moved as a group 
relative to the dead zone. Each module places a pair of silicon microstrip sensors back-to-back at a 
specified stereo angle with independent cooling and support. The sensors are read out
continuously at 40~MHz using the APV25 ASIC, developed for the CMS experiment at the
LHC.

Each half of the ECal consists of 221 PbWO$_4$ crystals arranged in rectangular formation. 
There are five rows with 46 
modules in each row except the row closest to the beam plane which has 37. The light from each crystal is read 
out by Avalanche Photodiode (APD) glued on the back surface of the crystal. 
Signals from the APDs were amplified using custom-made amplifier boards before being sent to the readout 
electronics.

@@

\subsection{SVT Digitization, Event Building and Data Transmission}
\label{sec:svt_daq}
The SVT data acquisition (DAQ) based on the Reconfigurable Cluster Element (RCE) and Cluster 
Interconnect (CI) concept developed at SLAC as generic building blocks for DAQ systems in high energy 
physics {\bf [need reference]}. 
The RCE is a generic computational building block, housed on a separate daughter card called 
Data Processing Module (DPM), that are realized on an ATCA front board called the Cluster On Board (COB), 
see Fig.~\ref{fig:cob}.
 \begin{figure}[]
\begin{center}
{\small
	\includegraphics[width=6cm]{figures/svt_daq_module}
	\caption{Picture of the SVT DAQ COB board with four data processing daughter cards visible.}
	\label{fig:cob}
}
\end{center}
\end{figure}
The Gen-I RCE used in the Test run consisted of a Virtex~5 FPGA with 1~GB of DDR3 RAM. 
The outputs of 12 SVT half-modules 
are digitized on the Rear-Transition-Module (RTM), a custom board on the back side of the ATCA crate, 
interfacing the HPS readout to the generic DAQ components on the COB. 
A pre-amplifier converts the APV25 differential current output to a different voltage output scaled to the sensitive 
range of a 14-bit ADC operating at the system clock of 41.667~MHz. 
The RTM is organized into four sections with each section supporting three SVT half-module hybrids (15 
APV25 channels). The RTM also includes a 4-channel Fiber Optic module and supporting logic which can be 
used to interface to the JLab trigger system supervisor.
Each section of the RTM, up to three hybrids, is input to a DPM which apply thresholds for data reduction and 
organizes the sample data into UDP datagrams. The DPM also hosts an I2C controller used to configure and 
monitor the APV25 chips on those three hybrids. A single ATCA crate with two COB cards was used, one 
supporting four DPMs and one supporting 3 DPMs and one DPM that is configured as the trigger and data 
transmission module. The two COB cards and their DPMs are interconnected with a 10Gb/s switch 
card~\cite{Larsen:2011zb} which 
also hosts two 1Gb/s Ethernet interfaces to the external SVT DAQ PC.

>>>>

%% main text
\section{Introduction}
\label{introduction}
One of the more striking anomalies in recent astrophysical observations is an excess of high energy 
electrons and positrons consistent with a large, unknown source of pair 
production~\cite{Adriani:2008zr, FermiLAT:2011ab}. The existence of a massive, hidden sector vector 
boson; a ``dark photon;" that mixes kinetically with the Standard Model photon and mediates 
annihilation or decay of dark matter particles is a theoretically attractive explanation of this cosmic-ray 
excess~\cite{Holdom:1985ag}\cite{ArkaniHamed:2008qn}.
\begin{center}
{\small
\begin{figure*}[ht]
    \includegraphics[width=0.8\textwidth]{figures/hps_testrun_rendering}
\caption{Rendering of the HPS Test apparatus installed on the beam line. SHOULD FIND A PIC WITH LOWER HEIGHT.}
\label{fig:testrundetector}
\end{figure*}
}
\end{center}

Heavy photons, aka "hidden sector" or "dark" photons, are particles with mass 10-1000 MeV which 
couple weakly to the SM photon, so can be radiated by electrons, and can decay into e+e- pairs, albeit 
at rates far below those of QED trident processes. They have been suggested by numerous Beyond 
Standard Model theories and as an explanation of the current muonic g-2 anomaly, and may couple 
directly to hidden sector particles that may constitute all or some of the dark matter. Current 
phenomenology highlights the 20-1000~MeV/c$^{2}$ mass range, and suggests that heavy photons 
couple to electric charge with strength $\epsilon$e, where 
$\epsilon$ is in the range of $10^{-3} -10^{-5}$. This range of parameters makes heavy photon searches 
viable at electron-positron colliders and in moderate energy fixed target electroproduction, and results 
from both approaches have already been reported. Fixed target experiments offer the highest sensitivity 
for heavy photons with mass $<1000$~MeV/c$^{2}$ ~\cite{Bjorken:2009mm}, but require large data 
sets and good mass resolution
to identify a small mass peak above the copious QED background. At small couplings, heavy photons 
become long-lived, so detection of a secondary decay vertex can reject the prompt QED background 
and boost experimental sensitivity.
The HPS experiment uses both invariant mass and vertex signatures by placing a $\approx 1$~m long 
silicon tracking and vertexing detector (10~cm) downstream  of a 0.25 \%X$_0$ tungsten) target and 
inside a dipole magnet~\cite{HPS_proposal_2010}. The experiment runs at high rates, exploiting the 
100\% duty cycle of the CEBAF accelerator to accumulate the need statistics. The trigger is provided by 
a fast electromagnetic calorimeter just downstream of the magnet.

The HPS Test run, a simplified version of the full HPS experiment, was proposed and approved at JLab 
as the first stage of the HPS experiment. Its
purposes included demonstrating that the apparatus and data acquisition systems are technically 
feasible and that the trigger rates and occupancies encountered in electron-beam running are as 
simulated. Given dedicated running time with electron beams, the HPS Test Run apparatus is capable of 
searching for heavy photons in unexplored regions of parameter space.    
The HPS Test apparatus was installed on April 19, 2012, and 
ran parasitically with the HDice experiment, using its photon beam, until May 18. The JLab run schedule 
precluded any dedicated electron beam running, but the HPS Test Run was allowed a short and 
valuable dedicated run with the photon beam.

This paper reviews the HPS Test Run apparatus; documents the successful performance of the trigger, 
data acquisition, silicon tracker, and ECal at the level assumed in calculating the physics reach of the 
experiment. Of particular importance, data from the dedicated photon beam running has been used to 
compare the measured trigger rates with those expected in simulation. The Test Run trigger rate is 
almost entirely due to photons which have converted to \ee pairs in a thin target upstream of HPS and it 
is sensitive to their multiple Coulomb scattering in the conversion target.  During electron beam running, 
the tails of the multiply Coulomb scattered primary beam are the dominant source of occupancy in the 
tracker and trigger rate in the electromagnetic calorimeter. 
The trigger rates in the Test Run are also sensitive to the tails of the multiple Coulomb
scattering distribution. Good agreement between the measured and simulated rates confirms
our understanding of multiple Coulomb scattering and 
the reliability of the background simulation used to benchmark the physics reach of the HPS experiment.

Searching for a small signal on a large background requires high statistics. This in turn requires the 
experiment operate with high beam currents, handle very high data rates, and tolerate high radiation 
levels. In particular, the following are required:
\begin{itemize}
\item large and uniform acceptance in the forward region close to the beam in order to catch boosted heavy photon decay products,
\item beam passage through the apparatus in vacuum, to eliminate direct interactions with the detector and minimize beam gas interactions, 
\item a flexible, redundant and efficient trigger selecting electron and positron pairs at rates up to 50kHz,
\item excellent track reconstruction efficiency for electrons and positrons,
\item good angular and momentum resolution to reconstruct invariant mass precisely,
\item excellent vertex resolution to discriminate displaced heavy photon decays from prompt QED backgrounds,
\item high rate electronics with excellent timing resolution to minimize out of time backgrounds,
\item data handling rates of 100~MB/s to permanent storage,
\item detector components that can survive and efficiently operate in a high radiation environment with 
local doses exceeding $100$~Mrad.
\end{itemize}

\section{Detector Overview}
\label{detector}
The HPS experiment will run in Hall B at JLab using the CEBAF electron beam, a 
499 MHz beam at energies ranging up to 11 GeV and currents up to 800 nA.  
The design of the experiment follows from the kinematics of \Aprime production which 
typically results in final state particles within a few degrees of the incoming beam, especially at low 
$m_{\textrm{A}^\prime}$. So detectors must be placed close to the beam. The intense electron beam 
and the degraded electrons produced in the target and spread out horizontally by the analyzing magnet 
constitute a "dead zone"  which must be completely avoided, so the apparatus is split vertically above 
and below the beam plane. The beam is transported in vacuum to minimize beam gas backgrounds. 
Even so, background rates in the detectors remain high, so high rate detectors,
a fast trigger, and excellent time tagging are required to minimize their impact. 
At masses below $2 m_\mu$ only decays to electrons are allowed, so the key trigger 
comes from a lead-tungstate calorimeter with 250~MHz readout.  
%Above the di-muon threshold, a 
%downstream muon detector completes the trigger acceptance and may allow triggering for pion final states 
%required for sensitivity near the $\rho$ mass.

A rendering of the HPS Test apparatus installed on the beam line is shown in 
Fig.~\ref{fig:testrundetector}.
Most technologies used in the HPS Test apparatus will be used in the HPS detector with some 
improvements to each of the systems; both from lessons learned during running of the HPS Test but 
also from less schedule and budget constraints.

An overview of the coverage, segmentation and performance of the HPS Test detector is given in 
Tab.~\ref{tab:detector-overview}. 
%\begin{multicols}{2}
\begin{center}
\begin{table*}[t]
{\small
\caption{Overview of the coverage, segmentation and performance of the HPS Test detector. }
\begin{tabular}{lccccccc}
\hline 
System & Coverage & \# channels & ADC & Time resolution & \# layers & Segmentation & Performance \\
 & (mrad) &  & (bit) & (ns) &  &  &  \\
\hline
SVT & $15<\theta_{y} < XXXX$ & 12800 & 14 & $\approx 2$~ns & 5  & $\approx 120~\mu$m $r-\phi$ & $\sigma_{d0,y}  \approx 100~\mu$m \\
& &  &  &  & (stereo layers) & $\approx 6~\mu$m $z$ & $\sigma_{d0,x} \approx 300~\mu$m \\
& &  &  &  &  &  & $\sigma_{d0,z}\approx 1$~mm \\
\hline
ECal & $15<\theta_{y} < XXXX$ & 442 & 12 & 4~ns & 1 & $1.3\times1.3$~cm$^2$  & $\sigma(E)/E \approx 4.5\%$ \\ 
 &  &  &  &  &  & $1.6\times1.6$~cm$^2$  &  \\ 
\hline
\end{tabular}
\label{tab:detector-overview}
}
\end{table*}
\end{center}
%\end{multicols}

The silicon tracking and vertexing detector for HPS Test, or SVT, operates in an existing vacuum 
chamber inside the pair spectrometer analyzing magnet in Hall B at JLab.  The design principles of the 
SVT are described in further detail in the HPS Test Run Proposal ~\cite{HPS_tPROP}. There are five 
measurement stations, or ``layers,'' placed immediately downstream of the target. Each layer comprises 
a pair of closely-spaced planes and each plane is responsible for measuring a single coordinate, or 
``view.'' Introduction of a stereo angle between the two planes of each layer provides three-dimensional 
tracking and vertexing throughout the acceptance of the detector with one redundant layer.

In order to accommodate the dead zone, the SVT is built in two halves that are mirror reflections of one 
another about the plane of the nominal electron beam.  Each half consists of five double-sided modules 
mounted on a support plate that provides services to the modules and allows them to be moved as a 
group relative to the dead zone. Each module places a pair of silicon microstrip sensors back-to-back at 
a specified stereo angle with independent cooling and support.

Each half of the electromagnetic calorimeter, or ECal, consists of 221 PbWO$_4$ crystals arranged in 
rectangular formation. There are five rows with 46 
modules in each row except the row closest to the beam plane which has 37. The light from each crystal 
is read out by Avalanche Photodiode (APD) glued on the back surface of the crystal. 
Signals from the APDs were amplified using custom-made amplifier boards before being sent to the 
readout electronics.

@@

\subsection{SVT Digitization, Event Building and Data Transmission}
\label{sec:svt_daq}
The SVT data acquisition (DAQ) based on the Reconfigurable Cluster Element (RCE) and Cluster 
Interconnect (CI) concept developed at SLAC as generic building blocks for DAQ systems in high energy 
physics . The RCE is a generic computational building block, housed on a separate daughter card 
called Data Processing Module (DPM), that are realized on an ATCA front board called the Cluster On 
Board (COB), see Fig.~\ref{fig:cob}.
 \begin{figure}[]
\begin{center}
{\small
	\includegraphics[width=6cm]{figures/svt_daq_module}
	\caption{Picture of the SVT DAQ COB board with four data processing daughter cards visible.}
	\label{fig:cob}
}
\end{center}
\end{figure}
The Gen-I RCE used in the Test run consisted of a Virtex~5 FPGA with 1~GB of DDR3 RAM. 
The outputs of 12 SVT half-modules 
are digitized on the Rear-Transition-Module (RTM), a custom board on the back side of the ATCA crate, 
interfacing the HPS readout to the generic DAQ components on the COB. 
A pre-amplifier converts the APV25 differential current output to a different voltage output scaled to the sensitive 
range of a 14-bit ADC operating at the system clock of 41.667~MHz. 
The RTM is organized into four sections with each section supporting three SVT half-module hybrids (15 
APV25 channels). The RTM also includes a 4-channel Fiber Optic module and supporting logic which can be 
used to interface to the JLab trigger system supervisor.
Each section of the RTM, up to three hybrids, is input to a DPM which apply thresholds for data reduction and 
organizes the sample data into UDP datagrams. The DPM also hosts an I2C controller used to configure and 
monitor the APV25 chips on those three hybrids. A single ATCA crate with two COB cards was used, one 
supporting four DPMs and one supporting 3 DPMs and one DPM that is configured as the trigger and data 
transmission module. The two COB cards and their DPMs are interconnected with a 10Gb/s switch 
card~\cite{Larsen:2011zb} which 
also hosts two 1Gb/s Ethernet interfaces to the external SVT DAQ PC.
